{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce

### 2013-07-10 manuscript/nm.tex
recent versions for submission

In the next job, Cloudbreak computes a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping windows. For the experiments reported in this paper we use a window size of 25bp. Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}). To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.

@@

We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}. DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR. We also attempted to evaluate MoDIL on the same data. All of these methods detect deletions. Insertions can be detected by BreakDancer, Pindel, and MoDIL. See Methods for details on how reads were aligned and each program was invoked. For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{Salignment_comparison}).

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../figures/chr2_sim_rocs_runtime.pdf}
\caption{Accuracy and runtime performance on a simulated data set. (a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2. Deletions from the Venter genome were randomly added to one or both haplotypes. Each point on a curve represents a different threshold on the confidence of predictions made by that tool. Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score. (b) ROC curves for insertion predictions. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
\label{chr2CombinedRoc}
\end{figure}

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Methods for a description of how we identified correct predictions. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY. For insertions, Cloudbreak's performance is similar to or slightly better than Pindel. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).

@@

It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../figures/NA18507_rocs_runtime.pdf}
\caption{Accuracy and performance on the 37X NA18507 sample. (a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}. (b) ROC curve for insertion prediction performance. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material). }
\label{NA18507CombinedRoc}
\end{figure}

>>>>

In the next job, Cloudbreak computes a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping windows. For the experiments reported in this paper we use a window size of 25bp. Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Fig. \ref{Sinsert_size_mixes}). To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.

@@

We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}. DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR. We also attempted to evaluate MoDIL on the same data. All of these methods detect deletions. Insertions can be detected by BreakDancer, Pindel, and MoDIL. See Methods for details on how reads were aligned and each program was invoked. For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Fig. \ref{Salignment_comparison}).

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../figures/chr2_rocs_runtimes_journal.pdf}
\caption{Accuracy and runtime performance on a simulated data set. (a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2. Deletions from the Venter genome were randomly added to one or both haplotypes. Each point on a curve represents a different threshold on the confidence of predictions made by that tool. Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score. (b) ROC curves for insertion predictions. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
\label{chr2CombinedRoc}
\end{figure}

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Methods for a description of how we identified correct predictions. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY. For insertions, Cloudbreak's performance is similar to or slightly better than Pindel. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig. \ref{Sscalability}).

@@

It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Fig. \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../figures/NA18507_rocs_runtimes_journal.pdf}
\caption{Accuracy and performance on the 37X NA18507 sample. (a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}. (b) ROC curve for insertion prediction performance. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material). }
\label{NA18507CombinedRoc}
\end{figure}
